---
title: "WK05"
date: "2025-10-30"
output:
  html_document:
    css: style.css
  pdf_document: default
---

#### **01 OSM Part** (ShapeFile)<br>
**Step One:** We are Going to Use *"Geofabrik"* Website and download the server\
**Step Two:** Navigate to Europe > UK > England > Greater London\
**Step Three:** Download <span style="color:Blue;">**greater-london-latest-free.shp.zip**</span>\
**Step Four:** Unzip File and Save it into DATA\

#### **02 London Boroughs** (Boundry > ShapeFile)\
**Step One:** Get the data go to the London data store.\
**Step Two:** Statistical GIS boundary files for London.\
**Step Three:** Download the <span style ="color:Red;">**statistical-gis-boundaries-london.zip**</span>\
**Step Four:** Unzip the Data and save it to DATA folder.\

#### **03 World cities** (Context > ShapeFile)\
**Step Five:** Download them from the ArcGIS HUB > Download > Shapefile.\

#### **04 UK Outline** (Context > Shapefile)\
**Step Six:** Download GADM data for the UK.\

#### **05 Airbnb:** (CSV)
**Step Seven:** Download the <span style="color:Green;">**listings.csv**</span> from the inside airbnb website for London.

#### **06 Goal:**
1. Compare Airbnb and hotels across London Boroughs
2. Join the Airbnbs to the MSOAs and sum (i.e. how many are in each MSOA)
3. Join the Hotels to the MSOAs and sum

```{r}
library(sf)
library(tidyverse)

# Reading OSM Data

OSM <- st_read("DATA/greater-london-251102-free.shp/gis_osm_pois_a_free_1.shp")%>%
  st_transform(., 27700) %>%
  #select hotels only
  filter(fclass == 'hotel')
```
```{r}
# Airbnb data
Airbnb <- read_csv("DATA/listings (2).csv") %>%
  # longitude is considered x value here, latitude is y
  st_as_sf(., coords = c("longitude", "latitude"), 
                   crs = 4326) %>%
    st_transform(., 27700)%>%
    #select entire places that are available all year
    filter(room_type == 'Entire home/apt' & availability_365 =='365')


#London Borough data is already in 277000
Londonborough <- st_read("DATA/statistical-gis-boundaries-london/ESRI/London_Borough_Excluding_MHW.shp")%>%
  st_transform(., 27700)

```
```{r}
# load world cities
Worldcities <- st_read("DATA/World_Cities/World_Cities.shp") %>%
  st_transform(., 27700)
```
```{r}
# filter based on key cities
Worldcities2 <- Worldcities %>%
  filter(CNTRY_NAME=='United Kingdom'&
           Worldcities$CITY_NAME=='Birmingham'|
           Worldcities$CITY_NAME=='London'|
           Worldcities$CITY_NAME=='Edinburgh')

# load UK outline
UK_outline <- st_read("DATA/gadm41_GBR_shp/gadm41_GBR_0.shp")%>%
  st_transform(., 27700)
```
Lets use st_join() joining based on a column, what we want to make is that each hotel will be a new row. each hotel will have also a borough column. 
```{r}
Hotels <-  st_join(Londonborough, OSM)

Airbnbs <-  st_join(Londonborough, Airbnb)

head(Hotels)
```

group by and summarize 
```{r}
Hotels_sum <- Hotels %>%
  # we need to list the columns we want to keep in the summarise
  group_by(., GSS_CODE, NAME)%>%
  # for each group count the number of rows and store in a column called accomodation count.
  summarise(`Accomodation count` = n())

Airbnb_sum <- Airbnbs %>%
  group_by(., GSS_CODE, NAME)%>%
  summarise(`Accomodation count` = n())
```

a spatial join in like a left join, the boroughs will always exist. 
```{r}
# the st_join data = 0 Hotels
Hotels %>%
  filter(NAME=="Barking and Dagenham")

# the group by and summarise data = 1 Hotel!
Hotels_sum %>%
  filter(NAME=="Barking and Dagenham")
```
using st feature to see if st_contains(A, B) - Does A (boroughs) contain B (OSM)\
Returns a list of boroughs (33 rows) and in each borough a list of hotels
```{r}
Hotels_example <-st_contains(Londonborough, OSM)

Hotels_example
```
we take the length of that list per borough 
```{r}
Accomodation_contained <- Londonborough%>%
  mutate(hotels_n = lengths(st_contains(., OSM)))%>%
  mutate(airbnbs_n = lengths(st_contains(., Airbnbs)))
```

```{r}
# the group by and summarise data = 1 Hotel!
Accomodation_contained %>%
  filter(NAME=="Barking and Dagenham")
```
```{r}
Hotels2 <-  st_join(OSM, Londonborough)

#London_data <-  st_join(London_borough_data1, London_borough_data2, join=st_equals)
```
**Mapping**
loading your shape object with tm_shape() (here sf object, but it could be a raster)
setting the map layer (e.g. polygons (tm_polygons()), symbols (tm_symbols()),raster (tm_raster()), text (tm_text()), lines (tm_lines()) etc).
applying a derived function (e.g. border (tm_borders()), fill (tm_fill()), markers (tm_markers()).

```{r}
remotes::install_github("r-tmap/tmap")

library(tmap)

# change to "view" for an interactive map
tmap_mode("plot")

```
```{r}
# set the shape
tm1 <- tm_shape(Accomodation_contained) + 
  # set the map layer
  # try changing this to tm_symbols()
  tm_polygons("hotels_n",
              col = "black", lwd=0.5, lty="dashed",)

# plot the map
tm1
```

just borders 
```{r}
tm_no_map_layer <- tm_shape(Accomodation_contained) + 
  # there is no tm_polygons() if we just want the map with no spatial data
  tm_borders(col = "darkblue")

tm_no_map_layer
```

Add information to the map 
```{r}
tmap_mode("plot")

# plot each map
tm1 <- tm_shape(Accomodation_contained) + 
  tm_polygons(fill ="hotels_n",
              col = "black", 
              lwd =0.5,
              lty="dashed",
              fill.chart = tm_chart_violin(),
              # above this was the same as before
              fill.scale = tm_scale_intervals(
                values="brewer.bu_pu",
                n=5,
                style="jenks"))

tm1
```

in the above code we have:<p>
- tm_scale_intervals() - used for numerical data<p>
- values - the values to use to style the map, here it’s blue from the colour brewer list.<p> There are a number of options here. See the next code chunk.<p>
- n - the number of breaks to have<p>
- style - how to decide where to make the breaks. We have a lot of choice for this. See the<p>
interval scale reference. Jenks looks for the natural breaks in the data and is often used.<p>
- fill_chart() produces a summary of the mapped data with a violin plot, there are many<p> options see the tmap chart vigette.<p>
More colors with brewer.bu_pu

```{r}
library(colorblindcheck)
library(cols4all)
#cols4all::c4a_gui()
print("Hello")
```

setting up breaks across two data sets 
```{r}
stats <- Accomodation_contained %>%
  st_drop_geometry() %>%
  dplyr::select(hotels_n, airbnbs_n) %>%  
  summarise(across(everything(), list(
    min = min,
    max = max,
    mean = mean,
    median = median,
    sd = sd
  )))
```

using the airbnb as the greatest range and apply them to both data sets
```{r}
library(classInt)
# Get Jenks breaks for 5 classes
breaks <- Accomodation_contained%>%
  st_drop_geometry()%>%
  #need a numeric vector not a dataframe or tibble
  pull(airbnbs_n) %>%             
  classIntervals(., n = 5, style = "jenks")
  
breaks$brks
```
Not let’s make the maps for Hotels and Airbnbs as per the original plan. Because we have our<p>
data of interest in two separate we have a few options:<p>

Make a tmap object for each dataset (i.e. the same as above but now for airbnb)<p>
Use a facet map - this allows small maps for each group (here the group is airbnb or hotel)<p>
Let’s set up the facets..<p>

```{r}
tm1 <- tm_shape(Accomodation_contained) +
    tm_polygons(
      fill = c("hotels_n", "airbnbs_n"),
      fill.scale = tm_scale_intervals(values = "brewer.blues", breaks=breaks$brks),
      # set the legend
      fill.legend = tm_legend(title="Accomodation count",
                              title.size=0.85,
                              size=0.8,
                              # plot outside of the main map
                              #explained below
                              position=tm_pos_out("right", 
                                                  "center",
                                                  pos.v = "center")),
      # all facets share the same legend
      fill.free = FALSE)+
  # make 2 rows for the facets
  tm_facets(nrow=2)

tm1
```

```{r}
# tm_pos_out() the arguments that follow set where to place the legend:
# right = in the area to the right side of the plot (each map is a 3x3 grid or possible area)
# center = in the center [adjusted to the center] of the right “column”
# pos.v = "center" = in the middle of the column (not top of bottom)
# The vertical (pos.v) and horizontal (pos.h) positions can also be set numerically where in relation to the top left hand corner of the object you are placing.

# 0 = left (pos.h) or bottom (pos.v) 1 = right (pos.h) or top (pos.v)
```

For now, we need to:

Change the labels (tm_layout())<p>
Add a compass (tm_compass())<p>
Add a scalebar (tm_scalebar())<p>
Add a credit to OSM/Airbnb (tm_credits())<p>

```{r}
# change headings or remove

tm2 <- tm1+
  tm_layout(panel.labels=c("Hotels", "Airbnb"),
            panel.show = TRUE
            #panel.label.bg.color = "transparent",
            )+
  
  tm_compass(type= "arrow",
            size=1.8, 
            position = tm_pos_out("right", 
                                  "center",
                                  pos.h= -0.05,
                                  pos.v =0.72))+
  tm_scalebar(text.size = 0.7,
              width=10,
              breaks=c(0,10,20),
              position = tm_pos_out("right",
                                    "center", 
                                    pos.h=0.075,
                                    pos.v = 0.68))+
  # we could use tm_credits to place sub-titles like (A) or (B)
  # on the map.
  tm_credits("(c) OpenStreetMap contrbutors and Air b n b",
             size=0.6,
             position = tm_pos_out("center",
                                   "bottom",
                                   pos.v = 1.5,
                                   pos.h=-0.02))

tm2
```

Saving the file into desktop
```{r}
tmap_save(tm2, "DATA/1_facet.png", width = 8, height = 6, units="in", dpi = 300)
```

```{r}
qtm(UK_outline)
```

```{r}
#UK box according to bboxfinder and using st to crop it
newbb <- c(xmin=-296000, ymin=5408, xmax=655696, ymax=1000000)

UK_outlinecrop <- UK_outline$geometry %>%
  st_crop(., newbb)
```
To add layers to an individual map we add another geometry, here we have out UK outline<p>
tm_polygons) then add our cities (tm_symbols) on top of the polygons.<p>

```{r}
tm3 <- tm_shape(UK_outlinecrop)+ 
  tm_polygons(col="darkslategray1")+
    tm_layout(frame=FALSE)+
  tm_shape(Worldcities2) +
    #adds the city points
    tm_symbols(shape=20,
               fill = "orange",
               #outline colour
               col="black",
               size=0.8)+
        #add the city labels, x and y move the label around the point
    tm_text("CITY_NAME", xmod=-1, ymod=-0.5)


tm3
```

add a visual grey bounding box 

```{r}
# st_bbox gives the bounding x and y coordinates 
Londonbb <- st_bbox(Accomodation_contained,
                    crs = st_crs(Accomodation_contained))%>%
# st_as_sfc coverts the coordinates to an sf object
  st_as_sfc()

tm4 <- tm3 +
  tm_shape(Londonbb)+ 
  tm_borders(col = "grey40", lwd = 3)+
    tm_layout(frame=FALSE,
            bg.color = "transparent")
```

arranging the map with tmap_arrange() - although it doesn't give alot of control
we will focus on:
tm_grob()
```{r}
inset <- tmap_grob(tm4, asp=1.1)

final_map<- tm2+
  tm_inset(inset,
             position = tm_pos_out("right", "center"))
tmap_save(final_map, "prac5_data/2_facet.png", width = 8, height = 6, units="in", dpi = 300)
```

grid package and relocation of the object<p>
make data long in order to make violin

```{r}
library(ggplot2)

accom_long <- Accomodation_contained %>%
  # drop geometry for plotting
  st_drop_geometry() %>%  
  pivot_longer(
    cols = c(hotels_n, airbnbs_n),
    names_to = "accom_type",
    values_to = "count"
  )
```

lets make a plot

```{r}
violin <- accom_long %>%
  ggplot(aes(x = accom_type, 
             # number
             y = count, 
             # hotels or airbnb 
             fill = accom_type)) +
  geom_violin(trim = FALSE, 
              color = "grey30", 
              alpha = 0.8) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "",
    x = "",
    y = "",
    fill = "black"
  ) +
  # change the labels from the column names to..
scale_x_discrete(labels = c("hotels_n" = "Hotels", 
                            "airbnbs_n" = "Airbnbs"))+
  theme_minimal(base_size = 13)+
  theme(
    # no legend
    legend.position = "none",
    # all text to black
    text = element_text(color = "black"),         # set all text to black
    axis.text = element_text(color = "black"),    # axis tick labels
    axis.title = element_text(color = "black"),   # axis titles (if used)
    plot.title = element_text(color = "black")    # title (if used)
  )
```

Here is the grid package, we:
<p>
plot our map (tm2)<p>
then “print” over the study area map and violin plot with the x and y controlling placement on the page.<p>
save the plot by<p>

```{r}
library(grid)

# Open PNG device
png("prac5_data/3_facet.png", width = 8, height = 6, units="in", res = 300)

tm2
print(tm4, vp = viewport(x=0.68, y= 0.25, width = 0.3, height = 0.35))
print(violin, vp = viewport(x=0.67, y= 0.83, width = 0.25, height = 0.35))


# Close device
dev.off()
```









