---
title: 'student number: 25127081'
output:
  html_document:
    css: styles.css
  word_document: default
  pdf_document: default
---

#### **Originality declaration**  

I, **Emily Deeb**, confirm that the work presented in this assessment is my own. Where information has been derived from other sources, I confirm that this has been indicated in the work.

---


#### **Part 01: Project Scope**


**1.1 Background:**

This section establishes the conceptual, policy, and contextual foundation for your analysis.  
Explain:

- The **topic area**, why it matters, and how it connects to broader urban, social, or demographic processes.
- Relevant **local or global events**, such as policy reforms, economic shocks, inequality trends, or external disruptions.
- Key **academic literature** explaining spatial processes (clustering, segregation, neighbourhood effects, environmental exposure, etc.).
- Governmental or institutional **policies** shaping the phenomenon (e.g., eviction moratoria, planning regulations, public health interventions).
- Any current **debates**, contradictions, or tensions shaping expectations.

This should read as a *narrative*, not a bullet list. You are framing the problem for a government client, just as instructed in lecture slides.

To reach the A band, ensure this background:  
- contextualises the study area,  
- identifies the scale of the issue,  
- uses at least a couple of relevant sources or policy references,  
- and prepares the reader for why spatial analysis is essential rather than optional.

**1.2 Research Question, Aim, and Hypothesis:**

- **Research Question**
  
  - Should be **spatial**, **specific**, **logical**, and **achievable** within the exam.
  - Must flow directly from the background, and demonstrate understanding of the broader system.
  - Needs to highlight the spatial mechanism you are investigating (e.g. clustering, differentiation, hotspots, spatial dependence).

- **Research Aim**
  
  - A broad statement of what the report intends to achieve, aligned with the RQ.
  - Should echo the policy or analytic need identified in the background.

- **Research Hypothesis**
  
  - A clear directional or relational prediction (e.g., “higher X is associated with higher Y”).  
  - Include the **null hypothesis** (no relationship / random distribution).
  - Make sure the hypothesis is measurable and testable with the methods you will use.

  A-band requires creativity + grounding in literature/policy; ensure your RQ demonstrates both.

**1.3 Data:**

**1.3.1 Data Source and Governance:**

Explain:

- Who collects the dataset  
- Why it exists  
- How frequently it is updated  
- Data accuracy, known biases, or reporting gaps  
- Spatial units represented (points, polygons, rasters)

Also briefly acknowledge:

- potential political or bureaucratic biases in collection  
- differences in reporting practices across space  
- and how these limitations might shape results.

**1.3.2 Data Structure and Variables:**

- Discuss:

  - Core variables relevant to your topic  
  - Unique identifiers needed for joins  
  - Temporal coverage and its relevance  
  - CRS information (and any inconsistencies)  
  - Geographic boundaries and whether they make sense for your analysis

- Spatial considerations:

  - **CRS**: Are layers misaligned? Do we need to project?  
  - **Boundary validity**: Are polygons meaningful for your RQ?  
  - **Physical constraints**: rivers, highways, borough sizes  
  - **MAUP implications**: is aggregation affecting interpretation?

- Explain the spatial joins you will conduct:

  - Attribute joins  
  - Spatial filters  
  - Intersection-based joins  
  - Point-in-polygon relationships  

- And justify the **unit of analysis**.

Add one sentence explaining how these decisions affect your ability to interpret spatial clustering, autocorrelation, or local hotspots reliably.

**1.3.3 Data Limitations:**

- Identify:

  - Missingness  
  - Data displacement  
  - Format inconsistencies  
  - Temporal mismatch  
  - Sampling bias  
  - Aggregation issues  
  - Null or NA values in key variables  

- Briefly explain how these limitations influence interpretation.

You must explicitly state how these limitations may propagate into later methods:  
e.g. KDE smoothing of sparse data, instability in LISA, sensitivity of weight matrices, or misclassification of hotspots.

**1.4 Assumptions & Methodological Limitations:**

Before analysis, state your methodological constraints:

- Assumptions about data accuracy  
- CRS choices and the effect on distances  
- Limitations specific to your chosen methods  
- Sensitivity to bandwidth, spatial weights, or scale  
- Ecological fallacy risk  
- MAUP influence  
- Time constraints restricting alternative methods (e.g., spatial regression)

Explain how these may influence conclusions.

To strengthen this:  
Include 1–2 sentences on **why your results should be interpreted cautiously**, and where uncertainty may arise in your pipeline.

**1.5 Analytical Approach:**

A clear, structured plan (this is required in the mark scheme):

1. **Load & explore data**  
   Summaries, initial plots, CRS checks, field types.

2. **Wrangling & joining**  
   Projection, clipping, filtering, standardising data (count/area).

3. **Point pattern analysis**  
   KDE, quadrat, Ripley’s K, DBSCAN (if possible).

4. **Spatial autocorrelation**  
   Moran’s I, Geary’s C, Getis–Ord G, LISA, G*.

5. **Weight matrix construction**  
   Justification: Queen vs KNN, row-standardised vs binary.

6. **Final outputs**  
   Clean maps with interpretive captions.

7. **Interpretation**  
   Linking patterns back to the RQ + literature.

8. **Reflection**  
   Biases, assumptions, limitations, potential extensions.

9. **Recommendations**  
   Spatial evidence → policy action.

Add one final sentence clarifying that the approach balances feasibility within exam constraints while still delivering robust and interpretable spatial insights.

---



#### **Part 02: Data Analysis**
#### **2.1 Load Libraries and read files:**
Explain briefly why each package is needed (wrangling, spatial ops, plotting, autocorrelation).  
Also remind the reader that **no work should be done from the console**, only inside the RMarkdown, as warned in exam guidance.

- Steps:
  - Load Librares for Reading, Wrangling, Spatial Ops, Plotting, Autocorrelation; needed for the entire report
  - Read CSV file and SHP file for Data Wraggling and Analysis $km^2$
  
#### **2.1.1** Load Libraries needed for the Report  
```{r loadlibraries}
# Load Required libraries
library(sf)              # modern spatial vector data (points/lines/polygons, CRS, spatial ops)
library(tidyverse)       # core data wrangling + plotting (dplyr, readr, tibble, etc.)
library(tmap)            # thematic mapping for sf/raster (static + interactive maps)
library(tmaptools)       # helper tools for tmap (basemaps, geocoding, bbox utilities)
library(janitor)         # quick data cleaning (clean_names, tabulations, etc.)
library(stringr)         # string/text handling (str_detect, str_replace, regex work)
library(here)            # safe, project-root-based file paths (here("data", "file.csv"))
library(raster)          # older raster package (still used, but being replaced by terra)
library(terra)           # modern fast raster + vector grids (rast, crop, mask, project)
library(classInt)        # create class breaks for choropleths (jenks, quantiles, etc.)
library(ggplot2)         # general plotting (histograms, violins, geom_sf, etc.)
library(grid)            # low-level graphics; used for custom layout and insets
library(spatstat)        # point pattern analysis (ppp, Ripley’s K, quadrats, KDE)
library(fpc)             # clustering methods; here for DBSCAN on point coords
library(ggspatial)       # spatial helpers for ggplot (basemaps, north arrow, scalebar)
library(colorblindcheck) # check palettes for colour-blind accessibility
library(cols4all)        # big palette collection + GUI for choosing colour schemes
library(plotly)
library(rJava)
library(OpenStreetMap)
```

#### **2.1.2** Read Files and organize according to type 
```{r readfiles}
# read CSv
HDI <- read_csv(here::here("homework_4", "HDR21-22_Composite_indices_complete_time_series.csv"),
                       locale = locale(encoding = "latin1"),
                       na = " ", skip=0)
# Read SHP File
shape <- st_read(here::here("homework_2",
                            "Washington_Counties_with_Natural_Shoreline___washsh_area",
                            "Washington_Counties_with_Natural_Shoreline___washsh_area.shp"))
# OR
world_cities <- sf::st_read(here("homework_3", "World_Cities", "World_cities.shp"))

# Check Layers of GPKG
layers <- st_layers(here("homework_3", "gadm41_ESP.gpkg"))
# Read GPKG Via Layer
spain <- sf::st_read(here("homework_3", "gadm41_ESP.gpkg"),
                      layer='ADM_ADM_0')

# Read Raster Layer
jan<-terra::rast(here("prac3_data", "wc2.1_5m_tavg_01.tif"))
# have a look at the raster layer jan
jan
```

#### **2.1.3** Set Coordinates then Transform and project to Local Coordinates
> Notes: Make Sure what units you are working on 

```{r CheckCRS}
# Check CRS 
print(Ausoutline)
# Do we have coordinates?
# Is it an SF or SP
```

#### **2.1.4** Transform from Simple Project to Simple Feature Project 
```{r SPSF}
# You might also encounter an `SP` object from the `sp` package. In this case i'd advise just transforming the `sp` object to `sf` and changing the projection

#From sf to sp
AusoutlineSP <- Ausoutline %>%
  as(., "Spatial")

#From sp to sf
AusoutlineSF <- AusoutlineSP %>%
  st_as_sf()
```

#### **2.1.5** Set CRS and then Project
> Notes: Mollweide projection saved to a new object. The Mollweide projection retains area proportions whilst compromising accuracy of angle and shape 

```{r SetTransform}
# Set Coordinates EPSG 4326 WGS84 (Global)
Ausoutline <- Ausoutline %>%
  st_set_crs(., 4326)

# Transform the Sf into the local coordinates https://epsg.io/3112
AusoutlinePROJECTED <- Ausoutline %>%
  st_transform(.,3112)
print(AusoutlinePROJECTED)

# Project Raster
pr1 <- pr1 %>%
  terra::project(., "+proj=longlat +datum=WGS84 +no_defs +type=crs")
plot(pr1)

# Set CSV Coordinates 
Airbnb <- Airbnb %>% 
  st_as_sf(., coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(., 27700)

```

#### **2.1.6** Simplify Geom

Check layer by plotting the geometry
```{r Check_Layer}
# lets plot and check layer
plot(Ausoutline$geom)
```

Simplify the Shp file using st_simplify function
Here the argument `dTolerance` controls the level of generalisation in the units of the map
> Notes: This should load quicker, but for 'publication' or 'best' analysis (i.e. not just demonstrating or testing) i'd recommend using the real file to ensure you don't simplify a potentially important variable. 

```{r Simplify_Geom}

AusoutSIMPLE <- Ausoutline %>%
  st_simplify(., dTolerance = 1000) %>%
  st_geometry()%>%
  plot()

```


#### **2.2 Clean and Process the Data:**

Overall Goal: 

This Process includes: 

- Cleaning and organize column types
- Checking and clean missing values
- Normalization (e.g., count per area)
- Making joins 
- Creating analysis-ready subsets

- why each step matters (e.g. CRS → distance accuracy; renaming columns → clarity; normalization → comparability)  
- and any issues encountered and resolved.

#### **2.2.1** Give me the Class for each column in order to evaluate if all values are the needed character
```{r FirstCheck}
# Check Class
Datatypelist <- report %>% 
  summarise_all(class) %>%
  pivot_longer(everything(), 
               names_to="All_variables", 
               values_to="Variable_class")
```

Rename Columns
```{r RenameColumns}
#changing column name using dplyr
df <- df %>%
  dplyr::rename(column1 = Data1, column2=Data2)

```

Turn to Numeric
```{r AcrossNumeric}
  mutate(across(where(is.numeric), round, 3))%>%
  # Round Ukdiff to 0
  mutate(across(UKdiff, round, 0))%>%
```

> *Summary: Found that Blabla*


#### **2.2.2** Make smaller data frame for needed analysis
```{r Cleanandfilter}
# Steps:
# (1) Clean_names smaller case + with underscore
# (2) Select only required columns
# (3) Filter values Removing Multiple for county, and look only at schools and science class as well as all grades

filterd_report <- report %>%
  janitor::clean_names() %>%
  select(county, organization_level, test_subject, count_met_standard, 
         count_of_students_expected_to_test, grade_level)%>%
  filter(county != "Multiple")%>%
  filter(organization_level == "School")%>%
  filter(test_subject == "Science")%>%
  filter(grade_level=="All Grades")


# filter based on key cities
Worldcities2 <- Worldcities %>%
  janitor::clean_names() %>%
  filter(CNTRY_NAME=='United Kingdom'&
           Worldcities$CITY_NAME=='Birmingham'|
           Worldcities$CITY_NAME=='London'|
           Worldcities$CITY_NAME=='Edinburgh')

```

```{r CleanandfilterCountryCode}
library(countrycode)
HDIcols<- HDI %>%
  clean_names()%>%
  select(iso3, country, gii_2019, gii_2010)%>%
  mutate(difference=gii_2019-gii_2010)%>%
  #not needed here as we can now use the country name...but see below
  mutate(iso_code=countrycode(country, origin = 'country.name', destination = 'iso2c'))%>%
  mutate(iso_code2=countrycode(iso3, origin ='iso3c', destination = 'iso2c'))

```

```{r Str_detect}
# code is saying to stringr look for the rows that
# contain the word E09 then filter and store again
LondonBoroughs<- LondonData %>% 
  filter(str_detect(`New code`, "^E09"))
```

```{r Select_Contains}
# A more Data Science way > selecting via certain words
# You use contains() when you don’t know all the column names, or the names are long, messy, inconsistent, or there are many that share a pattern.
LondonBoroughs_contains<-LondonBoroughs %>% 
  dplyr::select(contains("expectancy"), 
                contains("obese - 2011/12 to 2013/14"),
                contains("Ward name")) 
```

> *Summary: Found that Blabla*


#### **2.2.3** Normalize the Data using count per county and not raw counts as it is misleading as they need to be evaluated with a standard unit
```{r GroupandMutate_SUM+PERCENT}
# Steps:
# (1) Grouped by county for count
# (2) Omit NA we dont want no values as the create a point without value
# (3) Create column that is named total_county_met that is the sum of count_met_standard
# (4) Create column total country to test which is the sum of count of students expected to test
# (5) create column that calculates the percent of met vs to test

county_only <- filterd_report %>% 
  group_by(county)%>%
  na.omit()%>%
  summarise(total_county_met_standard=sum(count_met_standard),
            total_county_to_test=sum(count_of_students_expected_to_test))%>%
  mutate(percent_met_per_county=(total_county_met_standard/total_county_to_test)*100)

```

```{r GorupandSummarize}
# we need to list the columns we want to keep in the summarise
# for each group count the number of rows and store in a column called accomodation count.
# if the df doesnt have geom then you are fine
Hotels_sum <- Hotels %>%
  group_by(., GSS_CODE, NAME)%>%
  summarise(`Accomodation count` = n()) 
# For Count it better to use st_contain and not this 
```

> *Summary: Found that Blabla*


#### **2.2.4** Calculate Average in all of the STATE in order to compare later each COUNTY to the average state
```{r calcAVERAGE}
# Steps:
# (1) Create new value (pull) that's named state_average which is the mean value of percent_met_per_county

# Option 1
state_average <- county_only%>%
  summarise(state_average= mean(percent_met_per_county))%>%
  pull()

# Option 2
state_average2 <- mean(county_only$percent_met_per_county)

# Option 3
state_average3 <- county_only%>%
  select(percent_met_per_county)%>%
  summarise(state_average=mean(percent_met_per_county))%>%
  pull()

```
> *Summary: Found that Blabla*


#### **2.2.5** Count the counties above or below state Average in order to understand which counties are in relation to the state average
```{r calcABOVEANDBELLOW}
# Steps:
# (1) Create a dataframe that takes only the grouped county 
# (2) Assign to each county a difference column that calculates if the county is above or below state average (negative vale = below)
# (3) Round the difference_to_state column to the nearest whole number for accurate count
# (4) Create a column named above_below that evaluates when the value is below avg. (negative) 

county_only_above_below_state <- county_only %>%
  mutate(difference_to_state=(percent_met_per_county-state_average))%>%
    mutate(across(difference_to_state, ~ round(.x, 0)))%>%
    mutate(above_below = case_when(difference_to_state<0 ~ "below",
                                   difference_to_state>0 ~ "above",
                                   difference_to_state==0 ~ "equal"
                                   ))
```
> *Summary: Found that Blabla*


#### **2.2.6** left Join and st_join Cleaned Data frame to the county Polygon, for mapping and spatial analysis
**Method:** left Join 
> Notes: pure attribute join

```{r LeftJoin}
# Steps: 
# (1) Organize Columns names for easier match with the data frame (similar columns)
# (2) Left Join the data frame county_only_above_below_state by countylabe(SHP) and county(DF)
# (3) Remove dups using distinct
joined_data <- shape %>% 
  clean_names(.) %>%
  left_join(., 
            county_only_above_below_state,
            by = c("countylabe" = "county"))%>% 
distinct(.,lad15cd, .keep_all = TRUE)

# OR

LondonWardsMerged <- LondonWardsMerged %>% 
  left_join(WardData, 
            by = c("GSS_CODE" = "new_code"))%>%
  dplyr::distinct(GSS_CODE, .keep_all = T)%>%
  dplyr::select(GSS_CODE, ward_name, average_gcse_capped_point_scores_2014)


```
> *Summary: Found that Blabla*

**Method:** Spatial Join
> Notes: You have two sf objects. You want to join based on where things are, not what their IDs are.
Result: one row per hotel, with borough columns added.
> Problems: ⚠️ Problem for counts: after st_join(), a borough with zero hotels still appears once (just the polygon row), so group_by(borough) %>% summarise(n = n()) gives 1 instead of 0.

**Method:** St_contains and St_within or st_equals
> Notes: using st feature to see if st_contains(A, B) - Does A (boroughs) contain B (OSM)\
Returns a list of boroughs (33 rows) and in each borough a list of hotels

```{r st_join}
# Default = st_intersects
# st_join(points, polygons)

# Explicit predicate
# st_join(points, polygons, join = st_within)
# st_join(polygons, points, join = st_contains)

Hotels_with_borough <- st_join(OSM_hotels, Londonborough, join = st_within) # or st_contain or st_equals

Hotels_with_borough

```

Count lengths ie count within column 
```{r st_contain_length}

Accomodation_contained <- Londonborough%>%
  mutate(hotels_n = lengths(st_contains(., OSM)))%>%
  mutate(airbnbs_n = lengths(st_contains(., Airbnbs)))

```

```{r st_intersects}
intersect_indices <-st_intersects(BoroughMap, BluePlaques)
```

Or using drop Geom to count Stats and future histograms
```{r GroupandSummarize+DropGeom}

Hotels_stats <- Hotels_boro %>%
  st_drop_geometry() %>%                  # drop hotel geometry for summarise
  group_by(GSS_CODE, NAME) %>%
  summarise(
    n_hotels    = n(),                    # number of hotels *with data*
    avg_price   = mean(price, na.rm = TRUE),
    total_rooms = sum(rooms, na.rm = TRUE)
  )

```


#### **2.2.7** Plot Map for initial Spatial Understanding using the normalized count of Above and below State Average
```{r IntialPlot_above_below}
# Steps:
# (1) Set Mode for Plot
# (2) Create bounding box using joined data geom points
# (3) Set OSM as Background
# (4) Use boundary Shape + OSM Map + Joined Data Polygons
# (5) fill county by values above equal or below the state average
# (6) Create a readable map 
tmap_mode("plot")

# Optional OSM basemap from the county bbox
bbox_county <- joined_data %>% 
  st_bbox() %>% 
  tmaptools::read_osm(type = "osm", zoom = NULL)

tm_shape(bbox_county) +
  tm_rgb() +
  tm_shape(joined_data) +
  tm_polygons(
    fill = "above_below",
    fill.scale = tm_scale_categorical(
      values = "brewer.bu_pu"
    ),
    fill_alpha = 0.5
  ) +
  tm_compass(position = c("left", "bottom"), type = "arrow") +
  tm_scalebar(position = c("left", "bottom")) +
  tm_layout(
    main.title = "Counties above or below state average for science in all grades",
    main.title.position = "center"
  )
```

```{r IntialPlot_Dots}
#plot the blue plaques in the city
tmap_mode("plot")

tm_shape(BoroughMap) +
  tm_polygons(fill_alpha = 0.5)+
tm_shape(BluePlaques) +
  tm_dots(fill = "blue", size=0.1)

```
> *Summary: Found that Blabla the map is missing count per county just for understanding*

```{r}
summary(BluePlaques)
BluePlaquesSub <- BluePlaques[LondonWardsMerged,]

```


#### **2.2.8:** Dealing with Rasters (Extra)
```{r Raster}
# A better and more efficient way is to firstly list all the files stored within our directory with `dir_info()` from the `fs` pacakge, then use `dplyr` in conjunction with `str_detect()` from `stringr` to search for filenames containing `tif`. Finally just select the paths.

# look in our folder, find the files that end with .tif and 
library(fs)
dir_info("prac3_data/") 

# Locate tif
# Here, we're also using `pull()` from `dplyr` which is the same as the `$` often used to extract columns as in the next stage the input must be filenames as characters (nothing else like a column name).
listfiles<-dir_info("prac3_data/") %>%
  filter(str_detect(path, ".tif")) %>%
  dplyr::select(path)%>%
  pull()

#have a look at the file names 
listfiles

# Then load all of the data straight into a SpatRaster. A SpatRaster is a collection of raster layers with the same spatial extent and resolution.

worldclimtemp <- listfiles %>%
  terra::rast()
  
#have a look at the raster stack
worldclimtemp

# In the stack you'll notice that under dimensions there are 12 layers (nlyr). The stack has loaded the 12 months of average temperature data for us in order. 
# To access single layers within the stack:

# access the january layer
worldclimtemp[[1]]

# We can also rename our layers within the stack:

month <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
           "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

names(worldclimtemp) <- month

# Now to get data for just January use our new layer name

worldclimtemp$Jan

# Using a raster stack we can extract data with a single command!! For example let's make a dataframe of some sample sites --- Australian cities/towns.

site <- c("Brisbane", "Melbourne", "Perth", "Sydney", "Broome", "Darwin", "Orange", 
          "Bunbury", "Cairns", "Adelaide", "Gold Coast", "Canberra", "Newcastle", 
          "Wollongong", "Logan City" )
lon <- c(153.03, 144.96, 115.86, 151.21, 122.23, 130.84, 149.10, 115.64, 145.77, 
         138.6, 153.43, 149.13, 151.78, 150.89, 153.12)
lat <- c(-27.47, -37.91, -31.95, -33.87, 17.96, -12.46, -33.28, -33.33, -16.92, 
         -34.93, -28, -35.28, -32.93, -34.42, -27.64)
#Put all of this inforamtion into one list 
samples <- data.frame(site, lon, lat, row.names="site")
# Extract the data from the Rasterstack for all points 
AUcitytemp<- terra::extract(worldclimtemp, samples)

# Add the city names to the rows of AUcitytemp
# turn to a DataFrame

Aucitytemp2 <- AUcitytemp %>% 
  as_tibble()%>% 
  add_column(Site = site, .before = "Jan")


```

#### **2.2.9** Dealing with Rasters Crop and Mask
Check layer by plotting the geometry
```{r Check_Layer}
# lets plot and check layer
plot(Ausoutline$geom)
```

```{r Simplify_Geom}
# Simplify the Shp file using st_simplify function
# Here the argument `dTolerance` controls the level of generalisation in the units of the map
# Notes: This should load quicker, but for 'publication' or 'best' analysis (i.e. not just demonstrating or testing) i'd recommend using the real file to ensure you don't simplify a potentially important variable. 

AusoutSIMPLE <- Ausoutline %>%
  st_simplify(., dTolerance = 1000) %>%
  st_geometry()%>%
  plot()

```

```{r CropRaster}
# Set our map extent (where we want to clip the data to) to the outline of Australia then crop our WorldClim dataset to it.
# Notes: **HOWEVER**, we need to make sure that both of our layers are in the same coordinate reference system when we combine them

Austemp <- Ausoutline %>%
  # now crop our temp data to the extent
  terra::crop(worldclimtemp,.)

# plot the output
plot(Austemp)

```

```{r MaskRaster}
# Mask the Raster according to boundary
exactAus<-terra::mask(Austemp, Ausoutline)
```

#### **2.2.10** Pivot Longer + filter + mean
```{r Pivot_Longer+mean}
# As we did in practical 2, we need to put our variable (months) into a one column using ```pivot_longer()```. Here, we are saying select columns 1-12 (all the months) and place them in a new column called `Month` and their values in another called `Temp`
squishdata<-exactAusdf%>%
  pivot_longer(
  cols = 1:12,
  names_to = "Month",
  values_to = "Temp"
)

# Then subset the data, selecting two months using `filter()` from `dplyr`
twomonths <- squishdata %>%
  # | = OR
  filter(., Month=="Jan" | Month=="Jun")

# Get the mean for each month we selected, remember `group_by()` and `summarise()` from last week?
meantwomonths <- twomonths %>%
  group_by(Month) %>%
  summarise(mean=mean(Temp, na.rm=TRUE))

meantwomonths

```

```{r Pivot_Longer+Wide}
# 👉 Use pivot_longer when:
# You have many columns representing categories
# You want one row per category
# Graphing or mapping requires “tidy” data

# 👉 Use pivot_wider when:
# You want one row per spatial unit
# You want each variable/year to be its own column
# Shapefiles + joins work best this way

# Pivot Longer to reorganize the data
#convert the tibble into a tidy tibble
# take cols 4:11 and move them to tipping type
flytipping_long <- flytipping1 %>% 
  pivot_longer(
    cols = 4:11,
    names_to = "tipping_type",
    values_to = "count"
  )

# view the data
view(flytipping_long)

# Or use this
# an alternative which just pulls everything out into a single table
flytipping2 <- flytipping1[,1:4]

# Pivot Wider by making column for each year, 
# alter the data from pivot longer using the year
# and tipping Type
# pivot the tidy tibble into one that is suitable for mapping
flytipping_wide <- flytipping_long %>% 
  pivot_wider(
    id_cols = 1:2,
    names_from = c(year,tipping_type),
    names_sep = "_",
    values_from = count
  )

view(flytipping_wide)

# we are interested in a specific variable 
# each year on its own
widefly <- flytipping2 %>% 
  pivot_wider(
    names_from = year, 
    values_from = total_incidents)

```

#### **2.2.11** Organize Data according to order
mutate(Month = factor(..., levels = ...))
This converts the Month column into a factor with a specific order:
Jan → Feb → Mar → … → Dec
This matters because:
default factor order is alphabetical (Apr, Aug, Dec, …) → useless for time series
setting levels explicitly ensures:
✔️ plots show months in correct calendar order
✔️ summaries follow Jan–Dec
✔️ ggplot facets appear chronologically
If Month was already a factor, this reorders it.
If it was a character, it converts and orders it.

```{r reordersdf}
# Have you been getting an annoying error message about bin size and non-finate values: 
# dropped all the NAs with `drop_na()`
# made sure that the Month column has the levels specified, which will map in descending order (e.g. Jan, Feb, March..)
# selected a bin width of 5 and produced a faceted plot...

data_complete_cases <- squishdata %>%
  drop_na()%>% 
  mutate(Month = factor(Month, levels = c("Jan","Feb","Mar",
                                          "Apr","May","Jun",
                                          "Jul","Aug","Sep",
                                          "Oct","Nov","Dec")))
```

```{r reorder_desc}
arrange(desc(normalisedlifeepectancy))
```

#### **2.3 Summary of Key Initial Findings:**

- Summaries and Initial Observations:

  - BLABLA
  
- Early Anomalies:

  - BLABLA
  
- Plot Understanding:

  - BLABLA

- Next Steps According to Intial Findings:

  - Notes on spatial imbalance
  - Readings
  - Literature
  - Spatial Analysis
  
This section should transition naturally into the spatial methods that follow.

---

#### **Part 03: Spatial Analysis:**

#### **3.1 Spatial Joins & Spatial Descriptive Statistics:**

Confirm:

- All layers share a common CRS
- Correct geographic clipping
- Correct temporal filtering
- Correct spatial operations (st_join, st_intersection, spatial filters)
- Normalised variables
- Only relevant fields kept

Add a sentence on how these steps reduce analytical bias and ensure that downstream methods behave correctly.

#### **3.2 Descriptive Statistics:**
> Raster: When we extract data from a raster with a point or polygon it is termed **zonal statistics** in GIS - we want to get some statistics from the zone (point / polygon) for the area that is over the raster.

> General: Descriptive (or summary) statistics provide a summary of our data, often forming the base of quantitative analysis leading to inferential statistics which we use to make inferences about our data (e.g. judegements of the probability that the observed difference between two data sets is not by chance) 


#### **3.2.1 Method | ** Histogram
> A histogram lets us see the frequency of distribution of our data. This will vary based on the data that you have downloaded. 
> Requirments: 
  - The base `hist()` function needs it as numeric

**3.2.1.1** Create simple Histogram to view outlines Calculate the temprature in perth and the frequency 
> Notes: BLABLA

```{r histogram_simple}
hist(as.numeric(Perthtemp))

# There seems to be an outlier with this plot as we haven't removed the columns we don't need - look in the `Perthtemp` tibble and you will see an ID of 3 which is plotted above! We will solve this next
```
> *Summary: Found that Blabla*

Create a Histogram with selected breaks
> Notes: BLABLA

```{r histogram_better}

#define where you want the breaks in the historgram
userbreak<-c(8,10,12,14,16,18,20,22,24,26)

# remove the ID and site columns
Perthtemp <- Aucitytemp2 %>%
  filter(site=="Perth")

t<-Perthtemp %>%
 dplyr::select(Jan:Dec)

  hist((as.numeric(t)), 
     breaks=userbreak, 
     col="red", 
     main="Histogram of Perth Temperature", 
     xlab="Temperature", 
     ylab="Frequency")
  
```
> *Summary: Found that Blabla*

View histogram information
> Notes: BLABLA

```{r hint_info}
# Check out the histogram information R generated
# Here we have:
# breaks --- the cut off points for the bins (or bars), we just specified these
# counts --- the number of cells in each bin
# midpoints --- the middle value for each bin
# density --- the density of data per bin

histinfo <- as.numeric(t) %>%
  as.numeric()%>%
  hist(.)

histinfo
```
> *Summary: Found that Blabla*

**3.2.1.2** Histogram GGPLOT

Let's re-compute our histogram for Australia in March. We could just use hist like we have done before. We can either subset using the location (we know March is third in the raster).
> Notes: We need to make our raster into a data.frame to be compatible with ```ggplot2```, using a dataframe or tibble

```{r Filter_Subset}
# Option One
Perthtemp <- Aucitytemp2 %>%
  filter(exactAus=="March")

# Option Two
#subset using the known location of the raster
hist(exactAus[[3]], col="red", main ="March temperature")
```

Lets Make sure it a DataFrame for GGPLOT
```{r TurntoDataFrame}
exactAusdf <- exactAus %>%
  as.data.frame()
```

GGplot Histogram Temp vs Frequency + Mean Line (One Layer)
```{r ggplot_simple}
# set up the basic histogram
gghist <- ggplot(exactAusdf, 
                 aes(x=Mar)) + 
  geom_histogram(color="black", 
                 fill="white")+
  labs(title="Ggplot2 histogram of Australian March temperatures", 
       x="Temperature", 
       y="Frequency")
# add a vertical line to the hisogram showing mean tempearture
gghist + geom_vline(aes(xintercept=mean(Mar, 
                                        na.rm=TRUE)),
            color="blue", 
            linetype="dashed", 
            size=1)+
  theme(plot.title = element_text(hjust = 0.5))
```

GGplot Histogram Temp vs Frequency + Mean Line (Multi Layer) 
> Notes: See previous Part for Turning Data longer because it works best with Histogram

```{r ggplot_multi}
# Select the colour and fill based on the variable (which is our month). The intercept is the mean we just calculated, with the lines also based on the coloumn variable.
ggplot(twomonths, aes(x=Temp, color=Month, fill=Month)) +
  geom_histogram(position="identity", alpha=0.5)+
  geom_vline(data=meantwomonths, 
             aes(xintercept=mean, 
                 color=Month),
             linetype="dashed")+
  labs(title="Ggplot2 histogram of Australian Jan and Jun
       temperatures",
       x="Temperature",
       y="Frequency")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))
```

a more organized way is to do this is to use non-alphabatical order see part 02 for more detail
```{r ggpot_organized}
data_complete_cases <- squishdata %>%
  drop_na()%>% 
  mutate(Month = factor(Month, levels = c("Jan","Feb","Mar",
                                          "Apr","May","Jun",
                                          "Jul","Aug","Sep",
                                          "Oct","Nov","Dec")))
# Plot faceted histogram
ggplot(data_complete_cases, aes(x=Temp, na.rm=TRUE))+
  geom_histogram(color="black", binwidth = 5)+
  labs(title="Ggplot2 faceted histogram of Australian temperatures", 
       x="Temperature",
       y="Frequency")+
  facet_grid(Month ~ .)+
  theme(plot.title = element_text(hjust = 0.5))

```

How about an interactive histogram using ```plotly```
> Notes: This format of code where you set lots of varaibles then call them within a plot, package or fuction is sometihng you should become more familiar with as it's considerd good practice.

```{r Plotly}

# split the data for plotly based on month

jan <- squishdata %>%
  drop_na() %>%
  filter(., Month=="Jan")

jun <- squishdata %>%
  drop_na() %>%
  filter(., Month=="Jun")

# give axis titles
x <- list (title = "Temperature")
y <- list (title = "Frequency")

# set the bin width
xbinsno<-list(start=0, end=40, size = 2.5)

# plot the histogram calling all the variables we just set
ihist<-plot_ly(alpha = 0.6) %>%
        add_histogram(x = jan$Temp,
        xbins=xbinsno, name="January") %>%
        add_histogram(x = jun$Temp,
        xbins=xbinsno, name="June") %>% 
        layout(barmode = "overlay", xaxis=x, yaxis=y)

ihist

```

Let's change the pace a bit and do a quickfire of other descrptive statistics you might want to use
BOXPLOT
```{r BOXPLOT}
# mean per month
meanofall <- squishdata %>%
  group_by(Month) %>%
  summarise(mean = mean(Temp, na.rm=TRUE))

# print the top 1
head(meanofall, n=1)

# standard deviation per month
# how much the values spread out from the mean.
sdofall <- squishdata %>%
  group_by(Month) %>%
  summarize(sd = sd(Temp, na.rm=TRUE))

# maximum per month
maxofall <- squishdata %>%
  group_by(Month) %>%
  summarize(max = max(Temp, na.rm=TRUE))

# minimum per month
minofall <- squishdata %>%
  group_by(Month) %>%
  summarize(min = min(Temp, na.rm=TRUE))

# Interquartlie range per month
IQRofall <- squishdata %>%
  group_by(Month) %>%
  summarize(IQR = IQR(Temp, na.rm=TRUE))

# perhaps you want to store multiple outputs in one list..
lotsofstats <- squishdata %>%
  group_by(Month) %>%
  summarize(IQR = IQR(Temp, na.rm=TRUE), 
            max=max(Temp, na.rm=T))

# or you want to know the mean (or some other stat) 
#for the whole year as opposed to each month...

meanwholeyear=squishdata %>%
  summarize(meanyear = mean(Temp, na.rm=TRUE))
  
```

Quick Stats
```{r QuickStats}
stats <- Accomodation_contained %>%
  st_drop_geometry() %>%
  dplyr::select(hotels_n, airbnbs_n) %>%  
  summarise(across(everything(), list(
    min = min,
    max = max,
    mean = mean,
    median = median,
    sd = sd
  )))
```

Scatter Plot
```{r ScatterPlot}
#Fancier Plot
plot_ly(LondonBoroughs, 
        #data for x axis
        x = ~male_life_expectancy_2009_13, 
        #data for y axis
        y = ~percent_children_in_reception_year_who_are_obese_2011_12_to_2013_14, 
        #attribute to display when hovering 
        text = ~borough, 
        type = "scatter", 
        mode = "markers")

```

#### **3.3 Detecting Spatial Patterns:**
#### **3.3 Point Pattern Analysis (PPA):**

Explain:

- Whether points show randomness, clustering, or regularity  
- KDE for hotspot modelling  
- Quadrat for testing departures from CSR  
- Ripley’s K for distance-based clustering  
- DBSCAN for density-based segmentation  

Each method must be justified and its limitations acknowledged.

To strengthen this section:  
Include one or two “why this matters” statements tying PPA output back to policy or planning implications.

#### **3.3.1 Method |** PPP

Create an observation window for the analysis:
> Notes: we are using Harrow Window for analysis

```{r as_owin}
#now set a window as the borough boundary
window <- as.owin(Harrow)
plot(window)
```

For point pattern analysis, we need to create a point pattern (ppp) object.
> Notes: create a spatial object that confirms with package 

```{r sp_ppp}
#create a sp object
BluePlaquesSub<- BluePlaquesSub %>%
  as(., 'Spatial')
#create a ppp object from cooridnates from the window
BluePlaquesSub.ppp <- ppp(x=BluePlaquesSub@coords[,1],
                          y=BluePlaquesSub@coords[,2],
                          window=window)
```

Have a look at the new ppp object
> Notes: Print out the points that would be used for the analysis

```{r Plot_ppp}
BluePlaquesSub.ppp %>%
  plot(.,pch=16,cex=0.5, 
       main="Blue Plaques Harrow")

```
> *Summary: BLABLA*

#### **3.3.2 Method |** Kernel Density Estimation (KDE)

One way to summarise your point data is to plot the density of your points under a window called a 'Kernel'. The size and shape of the Kernel affects the density pattern produced, but it is very easy to produce a Kernel Density Estimation (KDE) map from a ppp object using the `density()` function.

> Notes: The sigma value sets the diameter of the Kernel (set in meters)

```{r KDE}
BluePlaquesSub.ppp %>%
  density(., sigma=500) %>%
  plot()
```
> *Summary: BLABLA* 


#### **3.3.3 Method |** Quadrant Analysis

We are interesting in knowing whether the distribution of points in our study area differs from 'complete spatial randomness' CSR
> Notes: Note, I wouldn't recommend doing a quadrat analysis in any real piece of analysis you conduct, but it is useful for starting to understand the Poisson distribution

```{r Quadrat_Analysis}
#First plot the points
plot(BluePlaquesSub.ppp,
     pch=16,
     cex=0.5, 
     main="Blue Plaques in Harrow")

#now count the points in that fall in a 6 x 6
#grid overlaid across the windowBluePlaquesSub.ppp2<-BluePlaquesSub.ppp %>%
BluePlaquesSub.ppp %>%
  quadratcount(.,nx = 6, ny = 6)%>%
    plot(., add=T, col="red")
  
```
> *Summary: In our case here, want to know whether or not there is any kind of spatial patterning associated with the Blue Plaques in areas of London. If you recall from the lecture, this means comparing our observed distribution of points with a statistically likely (Complete Spatial Random) distribution, based on the Poisson distribution.* 

Using the same `quadratcount()` function again (for the same sized grid) we can save the results into a table:
```{r Quadratcount}
#run the quadrat count
Qcount <- BluePlaquesSub.ppp %>%
  quadratcount(.,nx = 6, ny = 6) %>%
  as.data.frame() %>%
  dplyr::count(Var1=Freq)%>%
  dplyr::rename(Freqquadratcount=n)

```

Check the data type in the first column --- if it is factor, we will need to convert it to numeric
```{r Quadrat_class}
Qcount %>% 
  summarise_all(class)
```
> *Summary: OK, so we now have a frequency table --- next we need to calculate our expected values. The formula for calculating expected probabilities based on the Poisson distribution is.*

Calculate lambda for the Poisson Distribution Calculate expected Qcount
```{r calc_lambada}
# calc the Expected sum
sums <- Qcount %>%
  #calculate the total blue plaques (Var * Freq)
  mutate(total = Var1 * Freqquadratcount) %>%
  dplyr::summarise(across(everything(), sum))%>%
  dplyr::select(-Var1) 

lambda<- Qcount%>%
  #calculate lambda
  mutate(total = Var1 * Freqquadratcount)%>%
  dplyr::summarise(across(everything(), sum)) %>%
  mutate(lambda=total/Freqquadratcount) %>%
  dplyr::select(lambda)%>%
  pull(lambda)

```


Calculate expected using the Poisson formula from above $k$ is the number of blue plaques counted in a square and is found in the first column of our table
```{r calc_expected}

QCountTable <- Qcount %>%
  mutate(Pr=((lambda^Var1)*exp(-lambda))/factorial(Var1))%>%
  #now calculate the expected counts based on our total number of plaques
  #and save them to the table
  mutate(Expected= (round(Pr * sums$Freqquadratcount, 0)))

#Compare the frequency distributions of the observed and expected point patterns
plot(c(1,5),c(0,14), type="n",
xlab="Number of Blue Plaques (Red=Observed,Blue=Expected)", 
     ylab="Frequency of Occurances")
points(QCountTable$Freqquadratcount, 
       col="Red", 
       type="o", 
       lwd=3)
points(QCountTable$Expected, col="Blue", 
       type="o", 
       lwd=3)
```
> *Summary: Comparing the observed and expected frequencies for our quadrant counts, we can observe that they both have higher frequency counts at the lower end --- something reminiscent of a Poisson distribution. This could indicate that for this particular set of quadrants, our pattern is close to Complete Spatial Randomness (i.e. no clustering or dispersal of points). But how do we confirm this?*

#### **3.3.4 Method |** Chi Squared

This uses a Chi Squared test to compare the observed and expected frequencies for each quadrant 
A Chi-Squared test determines if there is an association between two categorical variables. The higher the Chi-Squared value, the greater the difference.
If the p-value of our Chi-Squared test is < 0.05, then we can reject a null hypothesis that says “there is no pattern - i.e. complete spatial randomness - in our data” (think of a null-hypothesis as the opposite of a hypothesis that says our data exhibit a pattern). What we need to look for is a value for p > 0.05. If our p-value is > 0.05 then this indicates that we have CSR and there is no pattern in our points. If it is < 0.05, this indicates that we do have clustering in our points. 

```{r Chi_squared}

teststats <- quadrat.test(BluePlaquesSub.ppp, nx = 6, ny = 6)

plot(BluePlaquesSub.ppp,pch=16,cex=0.5, main="Blue Plaques in Harrow")
plot(teststats, add=T, col = "red")
```

> *Summary: Here our p-value = 0.2594, implying complete spatial randomness.* 


#### **3.3.5 Method |** Ripley's K

Compare the observed distribution of points with the Poisson random model for a whole range of different distance radii. This is what Ripley's K function computes. 
> Notes: We can conduct a Ripley's K test on our data very simply with the `spatstat` package using the `kest()` function. 

```{r Ripley's K}
K <- BluePlaquesSub.ppp %>%
  Kest(., correction="border") %>%
  plot()

# we can also extract the plot into a dataframe
Kval <- as.data.frame(Kest(BluePlaquesSub.ppp, correction = "Ripley"))

```
> *Summary: Ripley’s K value for any circle radius $r$, The average density of points for the entire study region (of all locations) Multiplied by the sum of the distances $d_ij$ between all points within that search radius, Divided by the total number of points, n. I = 1 or 0 depending if $d_ij < r$*
> *The plot for K has a number of elements that are worth explaining. First, the Kpois(r) line in Red is the theoretical value of K for each distance window (r) under a Poisson assumption of Complete Spatial Randomness. The Black line is the estimated values of K accounting for the effects of the edge of the study area.*
> *Here, the correction specifies how points towards the edge are dealt with, in this case, border means that points towards the edge are ignored for the calculation but are included for the central points.*
> *Where the value of K falls above the line, the data appear to be clustered at that distance. Where the value of K is below the line, the data are dispersed. From the graph, we can see that up until distances of around 1300 metres, Blue Plaques appear to be clustered in Harrow, however, at around 1500 m, the distribution appears random and then dispersed between about 1600 and 2100 metres.* 

**Quadrat and Ripley's K analysis are useful exploratory techniques for telling us if we have spatial clusters present in our point data, but they are not able to tell us WHERE in our area of interest the clusters are occurring. To discover this we need to use alternative techniques.** 

#### **3.3.6 Method |** DBSCAN

We will now carry out a DBSCAN analysis of blue plaques in my borough to see if there are any clusters present.
DBSCAN requires you to input two parameters:
1. *Epsilon* - this is the radius within which the algorithm with search for clusters
2. *MinPts* - this is the minimum number of points that should be considered a cluster
> Notes: Based on the results of the Ripley's K analysis earlier, we can see that we are getting clustering up to a radius of around 1200m, with the largest bulge in the graph at around 700m. Therefore, 700m is probably a good place to start and we will begin by searching for clusters of at least 4 points

```{r DBSCAN_Simple}
#first check the coordinate reference system of the Harrow spatial polygon:
st_geometry(BoroughMap)

#first extract the points from the spatial points data frame
BluePlaquesSubPoints <- BluePlaquesSub %>%
  coordinates(.)%>%
  as.data.frame()

#now run the dbscan analysis
db <- BluePlaquesSubPoints %>%
  fpc::dbscan(.,eps = 700, MinPts = 4)

#now plot the results
plot(db, BluePlaquesSubPoints, main = "DBSCAN Output", frame = F)
plot(BoroughMap$geometry, add=T)

```
> *Summary: BLAVLA*

You could also use ```kNNdistplot()``` from the ```dbscan``` package to find a suitable eps value based on the 'knee' in the plot 
```{r KNN}
# used to find suitable eps value based on the knee in plot
# k is no of nearest neighbours used, use min points
library(dbscan)

BluePlaquesSubPoints%>%
  dbscan::kNNdistplot(.,k=4)
```
> *Summary: This plot shows for each point the average distance to the k neighbours, which are then plotted in ascending order. The knee is where this value (of distance to neighbours) increases.*
> *So the DBSCAN analysis shows that for these values of eps and MinPts there are three clusters in the area I am analysing.*


Our new db object contains lots of info including the cluster each set of point coordinates belongs to, whether the point is a seed point or a border point etc. We can get a summary by just calling the object
If you open up the object in the environment window in RStudio, you will also see the various slots in the object, including cluster

```{r DBSCAN_Better}
db$cluster

BluePlaquesSubPoints<- BluePlaquesSubPoints %>%
  mutate(dbcluster=db$cluster)

# Next we are going to create some convex hull polygons to wrap around the points in our clusters. This means we will create a smallest polygon that encloses the set of points of each cluster. We will
# convert this dataframe to an sf object
# make convex hulls around the clusters 
# plot

# convert the data frame to sf
BluePlaquesSubPoints_sf <- st_as_sf(BluePlaquesSubPoints, 
                                    coords = c("coords.x1", "coords.x2"), 
                                    crs = 27700)

# make the convex hulls around the cluster points
chull_polygons <- BluePlaquesSubPoints_sf %>%
  # remove any points not in a cluster
  filter(dbcluster>0)%>%
  group_by(dbcluster) %>%
  summarise(geometry = st_combine(geometry)) %>%  # combine points
  mutate(geometry = st_convex_hull(geometry)) %>% # convex hull
  st_as_sf()

# Use= `annotation_map_tile()` function from the ggspatial package
# `fill` is for the interior colour (e.g. polygons)
# `color` is for borders and points

library(ggspatial)

# Create the map
ggplot() +
  annotation_map_tile(zoom = 13) +
  geom_sf(data = BluePlaquesSubPoints_sf, aes(color=dbcluster), size = 3)+
  geom_sf(data = chull_polygons, aes(fill = dbcluster), 
          alpha = 0.8, 
          # remove polygon borders
          color = NA, 
          show.legend = FALSE) +

  theme_bw()

```


#### **3.4 Spatial Autocorrelation:**

- **Global Statistics:**

  - Moran’s I: overall clustering of values  
  - Geary’s C: more sensitive to neighbouring contrasts  
  - Getis–Ord G: global hotspots/coldspots

- **Local Statistics:**

  - Local Moran’s I (LISA)  
  - Local Getis–Ord G*

Explain interpretation categories: HH, LL, HL, LH.

Add one line:  
These local statistics allow us to spatially diagnose risk, vulnerability, or concentration in a way global indicators cannot.

- **Spatial Weight Matrix Rationale:**

  - **Queen Contiguity:**  
    - Border OR vertex adjacency  
    - Good for irregular polygons  
    - Uneven neighbour counts possible

  - **KNN:**  
    - Equal number of neighbours  
    - Distance-based, not boundary-based  
    - Can link across rivers/highways

  - **Row-Standardised Weights:**  
    - Each row sums to 1  
    - Corrects for unequal neighbour counts

  - **Binary Weights:**  
    - Equal treatment for all neighbours  
    - Can exaggerate influence of polygons with many neighbours

Add 2–3 sentences on *why your chosen matrix best reflects the spatial process* (e.g., diffusion vs adjacency vs distance decay).

**Analysing Spatial Autocorrelation with Moran's I, LISA and friends**

Patterns of spatially referenced continuous observations using various measures of spatial autocorrelation. Spatial autocorrelation is a measure of similarity between nearby data. 

> Notes: The measures of spatial autocorrelation that we will be using require continuous observations (counts of blue plaques, average GCSE scores, average incomes etc.) to be spatially referenced (i.e. attached to a spatial unit like a ward or a borough).
> To create a continuous observation from the blue plaques data we need to count all of the blue plaques that fall within each Ward in the City. Luckily, we can do this using the `st_intersects()` function from the `sf` package.
> `st_intersects()` will return a list of which points (plaques) are in each polygon (wards). So each row will be a polygon and the ID are the point IDs that intersect the polygon.
> In this case we don't want to use `st_join()` (which by default uses `st_intersects()`) as `st_join()` is like a left join and will retain all the data in on the left. So if a ward is empty, there will still be a row for that ward. So, we use `st_intersects()` on its own. We could also use other topological relationship functions such as `st_within()` instead...

```{r St_intersects2}
example<-st_intersects(LondonWardsMerged, BluePlaquesSub)

example

```
>*Summary: Here the polygon with the ID of 6 Kingston upon Thames - Coombe Hill has three plaques within it*

```{r st_join2}
check_example <- LondonWardsMerged%>%
  st_join(BluePlaquesSub)%>%
  filter(ward_name=="Kingston upon Thames - Coombe Hill")
```

Now we just take the length of each list per polygon and add this as new column...

```{r st_intersects_lengths}

points_sf_joined <- LondonWardsMerged%>%
  mutate(n = lengths(st_intersects(., BluePlaquesSub)))%>%
  janitor::clean_names()%>%
  #calculate area
  mutate(area=st_area(.))%>%
  #then density of the points per ward
  mutate(density=n/area)%>%
  #select density and some other variables 
  dplyr::select(density, ward_name, gss_code, n, average_gcse_capped_point_scores_2014)

```

How about a quick choropleth map to see how we are getting on...

```{r QuickMap}

points_sf_joined<- points_sf_joined %>%                    
  group_by(gss_code) %>%         
  summarise(density = first(density),
          wardname= first(ward_name),
          plaquecount= first(n))

tm_shape(points_sf_joined) +
    tm_polygons(
      fill = "density",
      fill.scale = tm_scale_intervals(
        values = "brewer.blues",
        style="jenks"),
      # set the legend
      fill.legend = tm_legend(title="Blue Plaque Density",
                              title.size=0.85,
                              size=0.8,
                              # plot outside of the main map
                              #explained below
                              position=tm_pos_out("right", 
                                                  "center",
                                                  pos.v = "center")))
```

> *Summary: So, from the map, it looks as though we might have some clustering of blue plaques in the centre of London so let's check this with Moran's I and some other statistics.* 

#### **3.4.1 Set Weight Matrix:**

Before being able to calculate Moran's I and any similar statistics, we need to first define a \(W_{ij}\) spatial weights matrix

First calculate the centroids of all Wards in London

```{r Weight_Matrix}
#First calculate the centroids of all Wards in London

coordsW <- points_sf_joined%>%
  st_centroid()%>%
  st_geometry()
  
plot(coordsW,axes=TRUE)

```


#### **3.4.2 Neighbor list:**
Now we need to generate a spatial weights matrix. We'll start with a simple binary matrix of queen's case neighbours (otherwise known as Contiguity edges corners). This method means that polygons with a shared  edge or a corner will be included in computations for the target polygon
A **spatial weight matrix represents the spatial element of our data**, this means we are trying to conceptualize and model how parts of the data are linked (or not linked) to each other spatially, using rules that we will set. 

> Notes: A note on spatial weight matrices If the features share a boundary they are **contiguous**, this can also be classed as only common boundaries --- a **rook** (like chess a rook can move forwards or side wards) or any point in common (e.g. corners / other boundaries) --- a **queen** (like chess a queen can move forwards, backwards or on a diagonal).
> **Alternatively** instead of using contiguous relationships you can use **distance** based relationships. This is frequently done with k nearest neighbours in which k is set to the closest observations. e.g. K=3 means the three closest observations. 
> In the first instance we must create a neighbours list --- which is a list of all the neighbours. To do so we will use `poly2nb()` with the argument `queen=T` saying we want a to use Queens case. Let's see a summary of the output

```{r Neighbours_list}
#create a neighbours list
LWard_nb <- points_sf_joined %>%
  poly2nb(., queen=T)

```
> Summary: The output is a neighbour list object hence why i've called the object `LWard_nb`. We can get some summary info and plot the matrix.

```{r SummaryLWNB}
summary(LWard_nb)
```
> Summary: Here it is telling us that the average number of neighbours is 5.88. Now let's plot the neighbours.

```{r PlotLWNB}
#plot them
plot(LWard_nb, st_geometry(coordsW), col="red")
#add a map underneath
plot(points_sf_joined$geometry, add=T)
```

#### **3.4.3 Matrix style:**

From the weights list we must now make a spatial weight matrix. This makes a matrix the size of the number of neighbourhoods with values indicating if the elements in the rows are a neighbour or not. The style of weight is especially important here:

* B is the basic binary coding (1/0)
* W is **row** standardised (sums over all links to n)
* C is **globally** standardised (sums over all links to n)
* U is equal to C divided by the number of neighbours (sums over all links to unity)
* S is the variance-stabilizing coding scheme proposed

To understand this further let's start with binary. 

```{r B_Style}
#create a spatial weights matrix from these weights
Lward.lw <- LWard_nb %>%
  nb2mat(., style="B")

sum(Lward.lw)
```
> *Summary: Summing the binary (1/0) shows that we have 3680 neighbours. Now if we were to do global standardisation this would mean dividing our 625 wards by the total number of neighbours meaning each spatial weight has a value of 0.169.*

Alternatively we can do **row** standardisation where 1 is divided by the sum of the number of neighbours in each row. For example, row 1 here sums to 6, meaning each weight would be 0.166 in row 1 only.

```{r SUMLward}
sum(Lward.lw[1,])
```
> *Summary: Standardisation permits comparable spatial parameters. But you also need to consider geographical features*

#### **3.4.4 Autocorrelation:**

Now we have defined our \(W_{ij}\) matrix, we can calculate the Moran's I and other associated statistics. However, Moran's I requires a spatial weight **list** type object as opposed to **matrix**, this is simply...

```{r Weight_Style}
Lward.lw <- LWard_nb %>%
  nb2listw(., style="C")
```

#### **3.4.5 Method |** Moran's I

Moran's I test tells us whether we have clustered values (close to 1) or dispersed values (close to -1). We will calculate for densities rather than raw values.

Here:

* Clustered refers to how similar values are to the neighbours. This could be *any value* - high or low. 
* Dispersed means how dissimilar values are to neighbours 
* Random values would produce a result of 0.

Our neighbours are set by the weight matrix. 

```{r MoranI}
I_LWard_Global_Density <- points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  moran.test(., Lward.lw)
# moran.test(., Lward.lw, zero.policy = TRUE)`

I_LWard_Global_Density
```
> *Summary: By defult all of our autocorrelation tests assume that we have spatailly continuous observations where all polygons have neighbours. However, depending on the weight matrix this might not be the case! An additional argument allows the function to permit spatial units with no neighbours*
> The Moran's I statistic = 0.67 (remember 1 = clustered, 0 = no pattern, -1 = dispersed) which shows that we have some distinctive clustering. But this could be clustering of any value.

* Interpretation: We have some clustering of  values (low or high) *across the entire study area*. 


#### **3.4.6 Method |** Geary's C

Geary's C as well..? This tells us whether **similar values or dissimilar values are clustering**. Here:

* C < 1 indicates that neighbours with similar values are clustered (compared to random chance)
* C >1  indicates that neighbours with dissimilar values are clustered (compared to random chance). 
* C = 0 indicates not autocorrelation


```{r GearyC}
C_LWard_Global_Density <- 
  points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  geary.test(., Lward.lw)

C_LWard_Global_Density
```

> *Summary: We should now have an idea about what the main differences between Moran's I and Geary's C are, and how to inptert the outputs. 

> **The Geary's C statistic = 0.41 (remember Geary's C falls between 0 and 2; 1 means no spatial autocorrelation, <1 - positive spatial autocorrelation or similar values clustering, >1 - negative spatial autocorreation or dissimilar values clustering), which shows that similar values are clustering**

> *Interpretation: We have some clustering of similar values *indicating neighbours are similar. 


#### Morans I Geary C You will notice that the **key difference is with the numerator**

| Measure       | What it compares                 |
| ------------- | -------------------------------- |
| **Moran’s I** | Value of unit vs **global mean** |
| **Geary’s C** | Value of unit vs **neighbour**   |


Key message:

- For Moran's I: 

  - we are looking at the product of deviaitions from the global mean (subtracting our central value $(z_i - \bar{z})$ and niehgbour $(z_j - \bar{z})$ from the global mean individually, then multiplying them).
  - If both **deviations are positive then this suggests clustering**

For Geary's C:

  - This changes to the difference between the central value and the neighbour, squared $(z_i - z_j)^2$
  - A **large difference means more dissimiarlity between the values**






#### **3.4.7 Method |** Getis Ord

Getis Ord General G...? This tells goes further and says if **high or low values are clustering**. Here:

* If G > expected = High values clustering, which could be termed a "hot spot"
* If G < expected = low values clustering, which could termed a "cold spot"
* If G = expected, then it is spatially random

> Notes: **Note** conceptaully Gertis Ord General G is again very similar to Moran's I and Geary's C but the focus is now on the *weighted product of values* - this means multiplying neighbours and weighting them with our weight matrix (in the numerator). **If a high value is near a high value it returns a larger product**.

```{r GetisOrd}
G_LWard_Global_Density <- 
  points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  globalG.test(., Lward.lw)

G_LWard_Global_Density
```
> *Summary: The General G statistic = G > expected, so high values are tending to cluster.
* Interpretation: We have clustering *of high values*.

 
#### **3.4.8 Method |** Local Moran's I 

We can now also calculate local versions of the Moran's I statistic (e.g. for each Ward) statistic to spatially see and map *where* we have clusters...
This iteraties over each individual polygon (or point), using just the neighbours for the spatial unit from our weight matrix.

Local Moran's I is:

- The difference between a value and the global mean, the difference between a neighbour and the global mean, weighted by the weight matrix
- This is standarised by the global variance $(m_2)$ to make sure values are comparable. 
- It returns several columns, of most interest is the Z score. A Z-score is **how many standard deviations a value is away (above or below) from the mean**. This allows us to state if our value is significantly different than expected value at this location considering the neighours. 
- We are comparing our value of Moran's I to that of an expected value (computed from a separate equation that uses the spatial weight matrix, and therefore considers the neighbouring values). 
- We are expecting our value of Moran's I to be in the middle of the distribution of the expected values. These expected values follow a normal distribution, with the middle part representing complete spatial randomness. This is **typically** between < -1.65 or > +1.65 standard deviations from the mean  
- The **null hypothesis** is always there is complete spatial randomness. A null hypothesis means:
- If our value is towards the tails of the distribution then it is unlikely that the value is completely spatially random and we can reject the null hypothesis...as it is not what we expect at this location.
- In the example where we use a z-score of >2.58 or <-2.58 we interpret this as: 2.58 or <-2.58 standard deviations away from the mean are significant at the 99% level...this means there is a <1% chance that autocorrelation is not present, indicating **positive autocorrelation** (this could be **high surrounded by high or low surrounded by low**). 
- Positive autocorrelation means a value at a location is similar to the neighbours. The values could be high or low.
- A low or negative Z score is indicative of spatial outlier (e.g. high next to low values)

* What we are comparing values to in Local Moran's I
* What the results mean 
* Why the results could be important

```{r localMoranI}
#use the localmoran function to generate I for each ward in the city

I_LWard_Local_count <- points_sf_joined %>%
  pull(plaquecount) %>%
  as.vector()%>%
  localmoran(., Lward.lw)%>%
  as_tibble()

I_LWard_Local_Density <- points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  localmoran(., Lward.lw)%>%
  as_tibble()

#what does the output (the localMoran object) look like?
slice_head(I_LWard_Local_Density, n=5)
```

> Summary: There are 5 columns of data. We want to copy some of the columns (the I score (column 1) and the z-score standard deviation (column 4)) back into the LondonWards spatialPolygonsDataframe

```{r CopyBack}
points_sf_joined <- points_sf_joined %>%
  mutate(plaque_count_I = as.numeric(I_LWard_Local_count$Ii))%>%
  mutate(plaque_count_Iz =as.numeric(I_LWard_Local_count$Z.Ii))%>%
  mutate(density_I =as.numeric(I_LWard_Local_Density$Ii))%>%
  mutate(density_Iz =as.numeric(I_LWard_Local_Density$Z.Ii))
```
 
We'll set the breaks manually based on the rule that data points >2.58 or <-2.58 standard deviations away from the mean are significant at the 99% level (<1% chance that autocorrelation not present); >1.96 - <2.58 or <-1.96 to >-2.58 standard deviations are significant at the 95% level (<5% change that autocorrelation not present). >1.65 = 90% etc.

```{r Breaks}
breaks1<-c(-1000,-2.58,-1.96,-1.65,1.65,1.96,2.58,1000)
```

Now create a new **diverging colour** brewer palette and reverse the order using `rev()` (reverse) so higher values correspond to red

```{r Moran_Color}
library(RColorBrewer)
MoranColours<- rev(brewer.pal(7, "RdGy"))
```

```{r Reverse_Color}

tm_shape(points_sf_joined) +
    tm_polygons(
      fill = "plaque_count_Iz",
      fill.scale = tm_scale_intervals(values = "brewer.br_bg",
                                      midpoint = "NA",
                                      style = "fixed",
                                      breaks=breaks1),
       fill.legend = tm_legend(title="Local Moran's I, Blue Plaques in London",
                              title.size=0.85,
                              size=0.8,
                              # plot outside of the main map
                              #explained below
                              position=tm_pos_out("right", 
                                                  "center",
                                                  pos.v = "center")))
                                      
```

> Summar: This map shows some areas in the centre of London that have relatively high scores, indicating areas with lots of blue plaques neighbouring other areas with lots of blue plaques.

#### **3.4.9 Method |** Local Getis Ord 

What about the Getis Ord statistic for hot and cold spots...

This is a very similar concept to Local Moran's I except it just returns a z-score...remember that a z-score shows **how many standard deviations a value (our value) is away (above or below) from the mean (of the expected values)**

**However**, in the case of Getis Ord \(G_{i}^{*}\) this is the local sum (of the neighbourhood. This means for each point $i$ sum all of the surrounding values based on the weight matrix) $\sum_{j} w_{ij} x_j$ compared to the sum of all features in the entire dataset $\sum_{j} x_j
$

In Moran's I this is just the value of the spatial unit (e.g. polygon of the ward) compared to the neighbouring units.

Here, to be significant (or a hot spot) we will have a high value surrounded by high values. The local sum of these values will be different to the expected sum (think of this as all the values in the area) then where this difference is large we can consider it to be not by chance...

The same z-score criteria then applies as before..

> The result of Getis Ord \(G_{i}^{*}\) analysis is an array of Z-scores, one for each pixel [or polygon], which is the number of standard deviations that the pixel [or polygon] and its neighbours are from the global mean. High Z-scores indicate more intense clustering of high pixel values, indicating hot spots. Low Z-scores indicate more intense clustering of low values, indicating cold spots. Individual pixels with high or low values by themselves might be interesting but not necessarily significant.

```{r GettisOrd}
Gi_LWard_Local_Density <- points_sf_joined %>%
  pull(density) %>%
  as.vector()%>%
  localG(., Lward.lw)

head(Gi_LWard_Local_Density)

```

Add the Getis Ord \(G_{i}^{*}\) data to the simple feature
```{r JoinGettis}
points_sf_joined <- points_sf_joined %>%
  mutate(density_G = as.numeric(Gi_LWard_Local_Density))
```

And map the outputs
```{r GettisMap}

tm_shape(points_sf_joined) +
    tm_polygons(
      fill = "density_G",
      fill.scale = tm_scale_intervals(values = "brewer.br_bg",
                                      midpoint = "NA",
                                      style = "fixed",
                                      breaks=breaks1),
       fill.legend = tm_legend(title="Gi*, Blue Plaques in London",
                              title.size=0.85,
                              size=0.8,
                              # plot outside of the main map
                              #explained below
                              position=tm_pos_out("right", 
                                                  "center",
                                                  pos.v = "center")))
```

#### Local Morans I VS GetisORD

| Statistic             | What is compared                               | Purpose                                                                                           |
| --------------------- | ---------------------------------------------- | ------------------------------------------------------------------------------------------------- |
| **Local Moran’s I**   | Deviation of a point from neibhouring values    | Identify clusters or spatial outliers (similar/dissimilar)                                        |
| **Moran’s I Z-score** | Observed Moran’s I vs expected under randomness | Measure statistical significance of spatial autocorrelation at each location; high values surrounded by high values or low values surrounded by low values
| **Getis-Ord \(G_i^*\)** | Local weighted sum vs global sum of all values | Identify hotspots (high values clustering) or coldspots (low values clustering)    


#### **3.4.10 Method |** add different Variable
> Summary: The local Moran's I and \(G_{i}^{*}\) statistics for wards clearly show that the density of blue plaques in the centre of the city exhibits strong (and positive) spatial autocorrelation.

```{r Check_variables}
#use head to see what other variables are in the data file

slice_head(points_sf_joined, n=2)
```

Or print out the class of each column like we did in week 2, although we need to drop the geometry. 

```{r Check_Class3}
Datatypelist <- LondonWardsMerged %>% 
  st_drop_geometry()%>%
  summarise_all(class) %>%
  pivot_longer(everything(), 
               names_to="All_variables", 
               values_to="Variable_class")

Datatypelist
```


```{r MergedGCSEMORAN}

I_LWard_Local_GCSE <- LondonWardsMerged %>%
  arrange(GSS_CODE)%>%
  pull(average_gcse_capped_point_scores_2014) %>%
  as.vector()%>%
  localmoran(., Lward.lw)%>%
  as_tibble()

points_sf_joined <- points_sf_joined %>%
  arrange(gss_code)%>%
  mutate(GCSE_LocIz = as.numeric(I_LWard_Local_GCSE$Z.Ii))

tm_shape(points_sf_joined) +
    tm_polygons(
      fill = "GCSE_LocIz",
      fill.scale = tm_scale_intervals(values = "brewer.br_bg",
                                      midpoint = "NA",
                                      style = "fixed",
                                      breaks=breaks1),
       fill.legend = tm_legend(title="Local Moran's I, GCSE Scores",
                              title.size=0.85,
                              size=0.8,
                              # plot outside of the main map
                              #explained below
                              position=tm_pos_out("right", 
                                                  "center",
                                                  pos.v = "center")))

```

Now the Gi* statistic to look at clusters of high and low scores and explain what the output map is showing and what other questions this can lead us to ask next week.....

```{r MergesGCSEGL}

G_LWard_Local_GCSE <- LondonWardsMerged %>%
  dplyr::arrange(GSS_CODE)%>%
  dplyr::pull(average_gcse_capped_point_scores_2014) %>%
  as.vector()%>%
  localG(., Lward.lw)

points_sf_joined <- points_sf_joined %>%
  dplyr::arrange(gss_code)%>%
  dplyr::mutate(GCSE_LocGiz = as.numeric(G_LWard_Local_GCSE))


tm_shape(points_sf_joined) +
    tm_polygons(
      fill = "GCSE_LocGiz",
      fill.scale = tm_scale_intervals(values = "brewer.br_bg",
                                      midpoint = "NA",
                                      style = "fixed",
                                      breaks=breaks1),
       fill.legend = tm_legend(title="Gi*, GCSE Scores",
                              title.size=0.85,
                              size=0.8,
                              # plot outside of the main map
                              #explained below
                              position=tm_pos_out("right", 
                                                  "center",
                                                  pos.v = "center")))

```

#### **Part 04: Visualization and Key Findings**

Create an OSM Map
```{r OSMmapWK02}
# Editable map, create a bounding box with 
# st_box() Function from the sf package
# st_bbox gives the bounding x and y coordinates 
tmaplondon <- BoroughDataMap %>%
    st_bbox(.) %>% 
  tmaptools::read_osm(., type = "esri", zoom = NULL) 

# note type="osm" gives error atm - issue raised on github: https://github.com/r-tmap/tmaptools/issues/41
  
# data first then + layer with with aestheics controlled
# through the arguments
# style -> how to divide the data into out color breaks
# Values -> the colour scheme to use
# Map style and values

tmap_mode("plot")

# add basemap as Red Green Blue raster
tm_shape(tmaplondon)+
  tm_rgb()+
  # add sf data
  tm_shape(BoroughDataMap) + 
  # add polygon layer
  tm_polygons(fill="rate_of_job_seekers_allowance_jsa_claimants_2015",
              fill.scale= tm_scale_intervals(values="brewer.bu_pu",
                                             style="jenks"),
              fill_alpha=0.5,
              fill.legend = tm_legend(title = "rate of claimants 2015", 
                                      size = 0.8))+
  tm_compass(type = "arrow", position = c("left", "bottom")) + 
  tm_scalebar(position = c("left", "bottom"))+
  tm_title("Job seekers' Allowance Claimants", 
           size = 2,
           position = c("center", "top"))


# To make the map interactive
BoroughDataMap84 <- BoroughDataMap %>%
st_transform(.,4326)

tmap_mode("view") # the tmap is set to view

# 
tm_shape(BoroughDataMap84) + 
  # add polygon layer
  tm_polygons(fill="rate_of_job_seekers_allowance_jsa_claimants_2015",
              fill.scale= tm_scale_intervals(values="brewer.bu_pu",
                                             style="jenks"),
              fill_alpha=0.5,
              fill.legend = tm_legend(title = "Job seekers' Allowance Claimants", 
                                      size = 0.8))+
  tm_basemap(server = "OpenStreetMap") +
  tm_compass(type = "arrow", position = c("left", "bottom")) + 
  tm_scalebar(position = c("left", "bottom"))+
  tm_title("Job seekers' Allowance Claimants", 
           size = 2,
           position = c("center", "top"))

## [view mode] Map component `tm_chart_none()` not supported in "view" mode.
## This message is displayed once per session.

# Merge our life expectancy 4 map with spatial data EW
# map our merge with tmap
Life_expectancy4map <- EW %>%
  inner_join(., 
             Life_expectancy4,
             by = c("lad15cd" = "new_code"))%>%
  distinct(.,lad15cd, 
           .keep_all = TRUE)


# change inner_join() to left_join
tmap_mode("plot")

tm_shape(tmaplondon)+
  # add basemap as Red Green Blue raster
  tm_rgb()+
  # add sf data
  tm_shape(Life_expectancy4map) + 
  # add polygon layer
  tm_polygons(fill="UKdiff",
              fill.scale= tm_scale_intervals(values="brewer.bu_pu",
                                             style="jenks"),
              fill_alpha=0.5,
              fill.legend = tm_legend(title = "Number of years", 
                                      size = 0.8))+
  tm_compass(type = "arrow", position = c("left", "bottom")) + 
  tm_scalebar(position = c("left", "bottom"))+
  tm_title("Difference in life expectancy", 
           size = 2,
           position = c("center", "top"))

```


Mapping With Tmap
> Requirments: 
- tm_shape() (here sf object, but it could be a raster)
- setting the map layer (e.g. polygons (tm_polygons()), symbols (tm_symbols()),raster (tm_raster()), text (tm_text()), lines (tm_lines()) etc).
- applying a derived function (e.g. border (tm_borders()), fill (tm_fill()), markers (tm_markers()).

Install
```{r tmap_install}

remotes::install_github("r-tmap/tmap")

library(tmap)

# change to "view" for an interactive map
tmap_mode("plot")

```

Polygons
```{r tmap_simple, fig.align='center'}

# set the shape
tm1 <- tm_shape(Accomodation_contained) + 
  # set the map layer
  # try changing this to tm_symbols()
  tm_polygons("hotels_n",
              col = "black", lwd=0.5, lty="dashed",)

# plot the map
tm1
```
> *Summary: BLABLA*

Polygons Just borders 
```{r tmap_borders, fig.align='center'}

tm_no_map_layer <- tm_shape(Accomodation_contained) + 
  # there is no tm_polygons() if we just want the map with no spatial data
  tm_borders(col = "darkblue")

tm_no_map_layer
```
> *Summary: BLABLA*

Polygons + Info
```{r tmap_simple_info, fig.align='center'}
# in the above code we have:
# (1) tm_scale_intervals() - used for numerical data
# (2) values - the values to use to style the map, here it’s blue from the colour brewer list.<p> There are a number of options here. See the next code chunk.
# (3) n - the number of breaks to have
# (4) style - how to decide where to make the breaks. We have a lot of choice for this. See the interval scale reference. Jenks looks for the natural breaks in the data and is often used.
# (5) fill_chart() produces a summary of the mapped data with a violin plot, there are many<p> options see the tmap chart vigette.
# More colors with brewer.bu_pu

tmap_mode("plot")

# plot each map
tm1 <- tm_shape(Accomodation_contained) + 
  tm_polygons(fill ="hotels_n",
              col = "black", 
              lwd =0.5,
              lty="dashed",
              fill.chart = tm_chart_violin(),
              # above this was the same as before
              fill.scale = tm_scale_intervals(
                values="brewer.bu_pu",
                n=5,
                style="jenks"))

tm1
```
> *Summary: BLABLA*

using the airbnb as the greatest range and apply them to both data sets
```{r Creating_Breaks}

library(classInt) # package that breaks maps

# Get Jenks breaks for 5 classes
breaks <- Accomodation_contained%>%
  st_drop_geometry()%>%
  #need a numeric vector not a dataframe or tibble
  pull(airbnbs_n) %>%             
  classIntervals(., n = 5, style = "jenks")
  
breaks$brks

```

Make a tmap object for each dataset
> Notes: Use a facet map - this allows small maps for each group (here the group is airbnb or hotel) Let’s set up the facets.

```{r tmap_2facets, fig.align='center'}

tm1 <- tm_shape(Accomodation_contained) +
    tm_polygons(
      fill = c("hotels_n", "airbnbs_n"),
      fill.scale = tm_scale_intervals(values = "brewer.blues", breaks=breaks$brks),
      # set the legend
      fill.legend = tm_legend(title="Accomodation count",
                              title.size=0.85,
                              size=0.8,
                              # plot outside of the main map
                              #explained below
                              position=tm_pos_out("right", 
                                                  "center",
                                                  pos.h = "center")),
      # all facets share the same legend
      fill.free = FALSE)+
  # make 2 rows for the facets
  tm_facets(nrow=2)

tm1
```
> *Summary: BLABLA*

```{r Generalrules}
# tm_pos_out() the arguments that follow set where to place the legend:
# right = in the area to the right side of the plot (each map is a 3x3 grid or possible area)
# center = in the center [adjusted to the center] of the right “column”
# pos.v = "center" = in the middle of the column (not top of bottom)
# The vertical (pos.v) and horizontal (pos.h) positions can also be set numerically where in relation to the top left hand corner of the object you are placing.

# 0 = left (pos.h) or bottom (pos.v) 1 = right (pos.h) or top (pos.v)
```

For now, we need to:
Change the labels (tm_layout())<p>
Add a compass (tm_compass())<p>
Add a scalebar (tm_scalebar())<p>
Add a credit to OSM/Airbnb (tm_credits())<p>

```{r tmap_2facets+symbols, fig.align='center'}
# change headings or remove

tm2 <- tm1+
  tm_layout(panel.labels=c("Hotels", "Airbnb"),
            panel.show = TRUE
            #panel.label.bg.color = "transparent",
            )+
  
  tm_compass(type= "arrow",
            size=1.8, 
            position = tm_pos_out("right", 
                                  "center",
                                  pos.h= -0.05,
                                  pos.v =0.65))+
  tm_scalebar(text.size = 0.7,
              width=10,
              breaks=c(0,10,20),
              position = tm_pos_out("right",
                                    "center", 
                                    pos.h=0.075,
                                    pos.v = 0.58))+
  # we could use tm_credits to place sub-titles like (A) or (B)
  # on the map.
  tm_credits("(c) OpenStreetMap contrbutors and Airbnb",
             size=0.6,
             position = tm_pos_out("center",
                                   "bottom",
                                   pos.v = 1.5,
                                   pos.h=-0.02))

tm2
```

Quick thematic Map
```{r qtm, fig.align='center'}
qtm(UK_outline)
```

Create Bounding Box
```{r CreateBox}
#UK box according to bboxfinder and using st to crop it
newbb <- c(xmin=-296000, ymin=5408, xmax=655696, ymax=1000000)

UK_outlinecrop <- UK_outline$geometry %>%
  st_crop(., newbb)
```

To add layers to an individual map we add another geometry, here we have out UK outline (tm_polygons) then add our cities (tm_symbols) on top of the polygons.
```{r tmap_outline+dots, fig.align='center'}
tm3 <- tm_shape(UK_outlinecrop)+ 
  tm_polygons(col="lightblue")+
    tm_layout(frame=FALSE)+
      tm_layout(
        fontfamily = "sans",
        fontface = "bold",
        frame = FALSE) +
  tm_shape(Worldcities2) +
    #adds the city points
    tm_symbols(shape=20,
               fill = "deeppink",
               #outline colour
               col="black",
               size=0.8)+
        #add the city labels, x and y move the label around the point
    tm_text("POP", xmod=0, ymod=-0.6,col = "slategray") + 
          tm_text ("CITY_NAME", xmod=0, ymod=0.6) + 
  tm_compass(type= "arrow",
            size=1.8, 
            position = tm_pos_in("right", 
                                  "bottom",
                                  pos.h= 0.95,
                                  pos.v =0.1)) + 

    tm_credits("(c) OpenStreetMap contrbutors and Airbnb",
             size=0.6,
             position = tm_pos_in("center",
                                   "bottom",
                                   pos.v = 0.02,
                                   pos.h=-0.50))


tm3
```

Add a visual grey bounding box 
```{r BoundingBox_visual}
# st_bbox gives the bounding x and y coordinates 
Londonbb <- st_bbox(Accomodation_contained,
                    crs = st_crs(Accomodation_contained))%>%
# st_as_sfc coverts the coordinates to an sf object
  st_as_sfc()

tm4 <- tm3 +
  tm_shape(Londonbb)+ 
  tm_borders(col = "navy", lwd = 1)+
    tm_layout(frame=FALSE,
            bg.color = "transparent")
tm4
```

arranging the map with tmap_arrange() - although it doesn't give alot of control
we will focus on:
tm_grob()
```{r tm_grob}
# Make a small inset map and place it inside the main map.
inset <- tmap_grob(tm4, asp=1.1)

final_map<- tm2+
  tm_inset(inset,
             position = tm_pos_out("right", "center"))
final_map

```

make data long in order to make violin
```{r violin_long}
accom_long <- Accomodation_contained %>%
  # drop geometry for plotting
  st_drop_geometry() %>%  
  pivot_longer(
    cols = c(hotels_n, airbnbs_n),
    names_to = "accom_type",
    values_to = "count"
  )
```

lets make a plot
```{r violin_plot}
violin <- accom_long %>%
  ggplot(aes(x = accom_type, 
             # number
             y = count, 
             # hotels or airbnb 
             fill = accom_type)) +
  geom_violin(trim = FALSE, 
              color = "grey30", 
              alpha = 0.8) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "",
    x = "",
    y = "",
    fill = "black"
  ) +
  # change the labels from the column names to..
scale_x_discrete(labels = c("hotels_n" = "Hotels", 
                            "airbnbs_n" = "Airbnbs"))+
  theme_minimal(base_size = 13)+
  theme(
    # no legend
    legend.position = "none",
    # all text to black
    text = element_text(color = "black"),         # set all text to black
    axis.text = element_text(color = "black"),    # axis tick labels
    axis.title = element_text(color = "black"),   # axis titles (if used)
    plot.title = element_text(color = "black")    # title (if used)
  )
violin
```

Here is the grid package, we:
plot our map (tm2)
then “print” over the study area map and violin plot with the x and y controlling placement on the page.
save the plot by
```{r print_all}
library(grid)

# Open PNG device
png("prac5_data/3_facet.png", width = 8, height = 6, units="in", res = 300)

tm2
print(tm4, vp = viewport(x=0.68, y= 0.25, width = 0.3, height = 0.35))
print(violin, vp = viewport(x=0.67, y= 0.83, width = 0.25, height = 0.35))


# Close device
# dev.off()
```

**4.1 Summary of Key Findings:**

- Direct answer to the RQ  
- Highlight largest spatial inequalities  
- Connect observed patterns to policy + literature  
- Identify unexpected or contradictory results  
- Explain spatial mechanisms (why patterns emerge where they do)

**4.2 Presentation of Visualisations:**

Ensure:

- Interpretive captions (not descriptive)  
- Clear legends, size, readability  
- Limited number of final outputs (quality > quantity)  
- Creative but appropriate cartographic choices  
- Consistency across all maps (colours, scales, boundaries)

**4.3 Evaluation Criteria:**

- Outputs are clear, concise, and well-designed.  
- Visualisations show creativity and effective communication.  
- Evidence of independent learning and thoughtful selection.  
- Quantity of charts is appropriate for the report length.

---

**Graph 01 - Evictions per year |**  
The analysis below shows summed evictions per year from Data Frame 3.0 (ev_year_counts).

> ##### **Summary Graph 01:**  
Eviction events decrease slightly from 2017 to 2019, with a dramatic drop in 2020. This decline directly correlated to the COVID-19 eviction freeze, which halted most executions across New York City.  
This matters to the report because we can see that 2020 alone cannot be treated as “normal” eviction activity.  
Using 2017–2019 provides needed context: the neighbourhood eviction footprint must be analysed throughout those years leading to 2020 to understand the pattern.

---

#### **Part 05: Reflection & Recommendations**

**5.1 Recap & Interpretation:**

- Bring together the full narrative  
- Compare to literature and theory  
- Discuss plausible mechanisms behind patterns  
- Identify contradictions and explain them  
- Assess whether your findings answer the RQ fully or partially

**5.2 Policy Recommendations:**

Should be:

- Targeted  
- Evidence-based  
- Feasible  
- Aware of political and spatial constraints  
- Linked clearly to your results  

Include at least one short paragraph explaining trade-offs or risks of misinterpretation.

**5.3 Study Limitations:**

Cover:

- Data quality  
- Methodological sensitivities  
- Spatial weights  
- MAUP + ecological fallacy  
- CRS, geometry distortions  
- Temporal or spatial omissions  
- Uncertainty in local statistics  
- Computational constraints

**5.4 Future Work:**

Suggest:

- Alternative datasets  
- Spatial regressions (SAR, SEM, GWR)  
- Different spatial units  
- More granular temporal analysis  
- Alternative weight matrices  
- Incorporating socio-economic covariates  
- Validating results with supplementary data sources

**5.5 Evaluation Criteria:**

- Discussion is engaging, reflective, and critically aware.  
- Clear linkage to relevant literature or policy debates.  
- Shows understanding of implications, caveats, and broader context.

---

#### **Part 06: Citations**

**[1]** Structural inequalities and housing instability  
Office of the New York State Comptroller. *New Yorkers in Need: The Housing Insecurity Crisis.* February 2024.  
https://www.osc.ny.gov/reports/new-yorkers-need-housing-insecurity-crisis
